---
title: "STAT 30030 Homework 3, Exercise 6"
author: "Robert Winter"
highlight-style: pygments
format: pdf
geometry:
      - top=30mm
      - left=30mm
toc: true
toc-title: Table of Contents
number-sections: true
editor: visual

# Wrap code chunk text
include-in-header:
  text: |
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

# Introduction

Suppose $T_1, \ldots, T_{10} \sim \mathrm{iid} \; \mathrm{Exp}(\lambda)$. Let $\bar{T}_n$ be the sample mean of $T_1, \ldots, T_{10}$. We consider two method of moments estimators for $\lambda$, generated from the first and second moments of any of the $T_i$'s, respectively:

$$
\hat{\lambda} = \frac{1}{\bar{T}_n} = \frac{n}{\sum_i T_i},
$$

and

$$
\hat{\lambda}_2=\sqrt{\frac{2n}{\sum_i T_i^2}}.
$$

We will empirically explore the properties of these two estimators.

# Part (i)

## Setup

```{r}
#| output: False
library(dplyr)
library(ggplot2)
```

## Simulation

First, we set the true value for our parameter $\lambda$ to be $3$.

```{r}
lambda = 3
```

We define a function `lambda_hat(Ts)` that computes the estimator $\hat{\lambda} = \frac{1}{\bar{T_n}} = \frac{n}{\sum_i{T_i}}$ on a sample of ten random variables $T_1, \ldots, T_{10} \sim \mathrm{iid} \; \mathrm{Exp}(\lambda)$.

```{r}
lambda_hat = function(Ts){
  Tn = 0
  for(Ti in Ts){
    Tn = Tn + Ti
  }
  Tn = Tn / 10
  lambda_hat = 1/Tn
  return(lambda_hat)
}
```

We define another function `lambda_2_hat(Ts)` that computes the estimator $\hat{\lambda}_2 = \sqrt{\frac{2n}{\sum_iT_i^2}}$ on a sample of ten random variables $T_1, \ldots, T_{10} \sim \mathrm{iid} \; \mathrm{Exp}(\lambda)$.

```{r}
lambda_2_hat = function(Ts){
  sum = 0
  for(Ti in Ts){
    sum = sum + Ti^2
  }
  lambda_2_hat = sqrt(2*10/sum)
  return(lambda_2_hat)
}
```

Finally, we generate 10,000 samples of ten random variables $T_1, \ldots, T_{10} \sim \mathrm{iid} \; \mathrm{Exp}(\lambda)$, and for each sample compute and store the estimates $\hat{\lambda}$ and $\hat{\lambda}_2$.

```{r}
data = tibble()
data = data %>%
  mutate(lambda_hat = 0) %>%
  mutate(lambda_2_hat = 0)

for(i in c(1:10000)){
  Ts = rexp(10, lambda)
  data = data %>% add_row(lambda_hat = lambda_hat(Ts),
                          lambda_2_hat = lambda_2_hat(Ts))
}
```

## Properties of $\hat{\lambda}$

```{r}
bias_lambda_hat = mean(data$lambda_hat) - lambda

variance_lambda_hat = var(data$lambda_hat)

mse_lambda_hat = variance_lambda_hat + bias_lambda_hat^2
```

Theoretically, the bias of $\hat{\lambda}$ should be $\frac{\lambda}{n-1} = \frac{3}{10-1} = \frac{1}{3} \approx 0.\overline{333}$. In our simulation, we estimated the bias of $\hat{\lambda}$ to be `r bias_lambda_hat`, which is roughly what we expected.

Moreover, theoretically, the variance of $\hat{\lambda}$ should be $\lambda^2\frac{n^2}{(n-1)^2(n-2)} = 3^2\frac{10^2}{(10-1)^2(10-2)} = \frac{900}{648} = \frac{25}{18} \approx 1.3\overline{88}$. In our simulation, we estimated the variance of $\hat{\lambda}$ to be `r variance_lambda_hat`, which is roughly what we expected.

Finally, theoretically, the MSE of $\hat{\lambda}$ should be $\mathrm{Var}(\hat{\lambda}) + \mathrm{bias}(\hat{\lambda})^2 = \frac{\lambda^2n^2}{(n-1)^2(n-2)}+\big(\frac{\lambda}{n-1}\big)^2 = \frac{25}{18}+\big(\frac{1}{3}\big)^2 = \frac{27}{18} = \frac{3}{2} = 1.5$. In our simulation, we estimated the variance of $\hat{\lambda}$ to be `r mse_lambda_hat`, which is roughly what we expected.

## Properties of $\hat{\lambda}_2$

```{r}
bias_lambda_2_hat = mean(data$lambda_2_hat) - lambda

variance_lambda_2_hat = var(data$lambda_2_hat)
# variance_lambda_2_hat = 0
# for(i in c(1:10000)){
#   variance_lambda_2_hat = variance_lambda_2_hat
#                           + (sum(data$lambda_2_hat)/nrow(data) 
#                           - data$lambda_2_hat[i])^2
# }
# variance_lambda_2_hat = variance_lambda_2_hat / (nrow(data)-1)

mse_lambda_2_hat = variance_lambda_2_hat + bias_lambda_2_hat^2
```

In our simulation, we find that the bias of $\hat{\lambda}_2$ is approximately `r bias_lambda_2_hat`, the variance of $\hat{\lambda}_2$ is approximately `r variance_lambda_2_hat`, and the MSE of $\hat{\lambda}_2$ is approximately `r mse_lambda_2_hat`.

## Comparison of $\hat{\lambda}$ and $\hat{\lambda}_2$

In our simulation analysis, we found that:

-   $\mathrm{bias}(\hat{\lambda}) < \mathrm{bias}(\hat{\lambda}_2)$;

-   $\mathrm{Var}(\hat{\lambda}) < \mathrm{Var}(\hat{\lambda}_2)$ ; and thus

-   $\mathrm{MSE}(\hat{\lambda}) < \mathrm{Var}(\hat{\lambda}_2)$.

Thus, $\hat{\lambda}$ is both less biased and more efficient than $\hat{\lambda}_2$, making $\hat{\lambda}$ the better estimator for $\lambda$, at least with respect to finite samples.

# Part (ii)

## Asymptotics of $\hat{\lambda}$

In @fig-lambda_hat, we observe that $\hat{\lambda}$ appears to be asymptotically distributed $\sim \mathcal{N}\big(\lambda, \frac{1}{10}\lambda^2\big)$, as we derived in class, with $n=10$.

```{r}
#| label: fig-lambda_hat
#| fig-cap: "Histogram of lambda_hat estimates and their distribution"
plot_lambda_hat = ggplot(data) +
  # Histogram
  geom_histogram(aes(x = lambda_hat, y = after_stat(density)), bins = 50,
                 fill = "lightgray", color = "black") +
  # normal density curve
  stat_function(fun = dnorm, args = list(mean = lambda, sd = sqrt(lambda^2/10)),
                color = "blue", linewidth = 1.2) +
  # aesthetics
  theme(panel.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted")) +
  # labels
  labs(x = bquote(hat(lambda)), y = "Density", title = "Method of Moments (First Moment) Parameter Estimates") +
  theme(plot.title = element_text(hjust=0.5))

plot_lambda_hat
```

## Asymptotics of $\hat{\lambda}_2$

In @fig-lambda_2_hat, we observe that $\hat{\lambda}_2$ appears to be asymptotically distributed $\sim \mathcal{N}\big(\lambda, \frac{1}{8}\lambda^2\big)$, as we derived in Exercise\~5(iii), with $n=10$.

```{r}
#| label: fig-lambda_2_hat
#| fig-cap: "Histogram of lambda_2_hat estimates and their distribution"
plot_lambda_2_hat = ggplot(data) +
  # histogram
  geom_histogram(aes(x = lambda_2_hat, y = after_stat(density)), bins = 50,
                 fill = "lightgray", color = "black") +
  # normal density curve
  stat_function(fun = dnorm, args = list(mean = lambda, sd = sqrt(lambda^2*5/40)),
                color = "blue", linewidth = 1.2) +
  # aesthetics
  theme(panel.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid.major = element_line(color = "gray", linetype = "dotted")) +
  # labels
  labs(x = bquote(hat(lambda)[2]), y = "Density", title = "Method of Moments (Second Moment) Parameter Estimates") +
  theme(plot.title = element_text(hjust=0.5))

plot_lambda_2_hat
```
